---
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-",
  fig.width = 8,
  fig.height = 5
)
set.seed(1)
```

# swjm: Stagewise Variable Selection for Joint Models of Semi-Competing Risks

## Overview

`swjm` implements stagewise (forward-stepwise) variable selection for joint
models of recurrent events and terminal events (semi-competing risks). Two
model frameworks are supported:

| Model | Type | `alpha` (first *p*) | `beta` (second *p*) |
|-------|------|---------------------|----------------------|
| **JFM** | Cox frailty | recurrence/readmission | terminal/death |
| **JSCM** | Scale-change (AFT) | recurrence/readmission | terminal/death |

Three penalty types are available: cooperative lasso (`"coop"`), lasso
(`"lasso"`), and group lasso (`"group"`). The cooperative lasso encourages
shared support between the readmission and death coefficient vectors.

## Installation

```{r install, eval = FALSE}
# From the package directory:
devtools::install("swjm")
```

## Data format

All functions expect a data frame with columns `id`, `t.start`, `t.stop`,
`event` (1 = readmission, 0 = terminal/censoring row), `status` (1 = death,
0 = alive/censored), and covariate columns `x1, ..., xp`.

---

## Workflow

### 1. Simulate data

```{r simulate}
library(swjm)

# Joint Frailty Model — scenario 1
set.seed(42)
dat_jfm  <- generate_data(n = 200, p = 100, scenario = 1, model = "jfm")
Data_jfm <- dat_jfm$data

cat(sprintf(
  "JFM data: n = %d subjects, %d readmission events, %d deaths\n",
  length(unique(Data_jfm$id)),
  sum(Data_jfm$event),
  sum(Data_jfm$status)
))

# JSCM — same scenario
set.seed(42)
dat_jscm  <- generate_data(n = 200, p = 100, scenario = 1, model = "jscm")
Data_jscm <- dat_jscm$data
```

### 2. Fit the stagewise regularization path

```{r stagewise}
fit_jfm <- stagewise_fit(Data_jfm, model = "jfm", penalty = "coop")
fit_jfm
```

The fit object now exposes `alpha` (readmission) and `beta` (death) as
separate *p* × (*k*+1) matrices — one column per stagewise step — in addition
to the combined `theta` matrix (2*p* × (*k*+1)):

```{r path-matrices}
p <- 100
cat("alpha matrix:", nrow(fit_jfm$alpha), "rows x", ncol(fit_jfm$alpha),
    "columns\n")
cat("beta  matrix:", nrow(fit_jfm$beta),  "rows x", ncol(fit_jfm$beta),
    "columns\n")

# Nonzero readmission coefficients at the final step
k_final <- ncol(fit_jfm$alpha)
cat("Nonzero alpha entries at final step:\n")
a_final <- round(fit_jfm$alpha[, k_final], 4)
print(a_final[a_final != 0])
```

`summary()` shows a compact table of path-end coefficients with variable type
(shared, readmission-only, or death-only):

```{r path-summary}
summary(fit_jfm)
```

### 3. Plot the coefficient path

`plot()` produces a glmnet-style coefficient trajectory plot with the number
of active variables on the top axis:

```{r plot-path, fig.height = 7}
plot(fit_jfm)
```

### 4. Cross-validation to select the tuning parameter

`cv_stagewise()` evaluates a cross-fitted estimating-equation score norm over
a grid of lambda values.

```{r cv, cache = TRUE}
lambda_path <- fit_jfm$lambda
dec_idx     <- swjm:::extract_decreasing_indices(lambda_path)
lambda_seq  <- lambda_path[dec_idx]

cv_jfm <- cv_stagewise(
  Data_jfm, model = "jfm", penalty = "coop",
  lambda_seq = lambda_seq, K = 3L
)
cv_jfm
```

The CV object stores `alpha` and `beta` at the optimal lambda, plus
`n_active_alpha`, `n_active_beta`, and `n_active` variable-count vectors
across the full lambda grid.

### 5. Plot the cross-validation results

```{r plot-cv}
plot(cv_jfm)
```

### 6. Summarize the chosen model

```{r coef-summary}
# summary() shows a formatted table of selected coefficients
summary(cv_jfm)
```

Direct access to selected coefficients:

```{r direct-coef}
cat("Selected readmission (alpha) variables:", which(cv_jfm$alpha != 0), "\n")
cat("Selected death (beta) variables:       ", which(cv_jfm$beta  != 0), "\n")

# coef() returns the combined numeric vector c(alpha, beta) for compatibility
theta_best <- coef(cv_jfm)
```

### 7. Baseline hazard

`baseline_hazard()` evaluates the cumulative baseline hazards at any desired
time points (Breslow for JFM; Nelson-Aalen on the accelerated scale for JSCM):

```{r baseline}
bh <- baseline_hazard(cv_jfm, times = c(0.5, 1.0, 2.0, 4.0, 6.0))
print(bh)
```

### 8. Predict survival curves

`predict()` computes subject-specific readmission-free and death-free survival
curves, together with the per-predictor contributions to each linear
predictor:

```{r predict}
# Three hypothetical new subjects (covariate vectors of length p = 100)
set.seed(7)
newz <- matrix(rnorm(300), nrow = 3, ncol = 100)
colnames(newz) <- paste0("x", 1:100)

pred <- predict(cv_jfm, newdata = newz)
pred

# S_re: readmission-free survival — first 10 time points
round(pred$S_re[, 1:10], 3)

# Predictor contributions for subject 1 — show only nonzero entries
contrib1 <- round(pred$contrib_re[1, ], 3)
contrib1[contrib1 != 0]
```

### 9. Plot survival curves and predictor contributions

`plot()` on a `swjm_pred` object draws four panels: survival curves for both
processes (all subjects in grey, highlighted subject in color) plus bar
charts of predictor contributions for the selected subject.

```{r plot-pred, fig.height = 7}
plot(pred, which_subject = 1)
```

---

### 10. JSCM workflow (cross-validation + survival prediction)

The same workflow applies to the JSCM.  `baseline_hazard()` and `predict()`
now work for both models: for JSCM, survival curves are estimated via a
Nelson-Aalen baseline on the accelerated time scale.  In the AFT
interpretation, $e^{\hat\alpha^\top z_i}$ is the time-acceleration factor for
subject $i$: greater than 1 means events happen sooner, less than 1 means
later.

```{r cv-jscm, cache = TRUE}
fit_jscm <- stagewise_fit(Data_jscm, model = "jscm", penalty = "coop")

lambda_path_jscm <- fit_jscm$lambda
dec_idx_jscm     <- swjm:::extract_decreasing_indices(lambda_path_jscm)
lambda_seq_jscm  <- lambda_path_jscm[dec_idx_jscm]

cv_jscm <- cv_stagewise(
  Data_jscm, model = "jscm", penalty = "coop",
  lambda_seq = lambda_seq_jscm, K = 3L
)
summary(cv_jscm)
```

```{r predict-jscm}
set.seed(7)
newz_jscm <- matrix(runif(600, -1, 1), nrow = 3, ncol = 100)

pred_jscm <- predict(cv_jscm, newdata = newz_jscm)

# Total time-acceleration factor per subject (recurrence process)
cat("Recurrence time-acceleration factors:\n")
print(round(pred_jscm$time_accel_re, 3))
```

`plot()` draws the same four-panel layout as for JFM: survival curves for
both processes plus bar charts of log time-acceleration contributions.

```{r plot-pred-jscm, fig.height = 8}
plot(pred_jscm, which_subject = 1)
```

---

### 11. Other penalties

```{r other-penalties, eval = FALSE}
fit_lasso <- stagewise_fit(Data_jfm, model = "jfm", penalty = "lasso")
cv_lasso  <- cv_stagewise(Data_jfm, model = "jfm", penalty = "lasso", K = 3L)
summary(cv_lasso)

fit_group <- stagewise_fit(Data_jfm, model = "jfm", penalty = "group")
cv_group  <- cv_stagewise(Data_jfm, model = "jfm", penalty = "group", K = 3L)
summary(cv_group)
```

---

## 12. Model evaluation

### 12.1 Coefficient recovery

Compare CV-optimal estimates to the true generating coefficients.
Variables that are truly nonzero or were selected are shown; all others
were correctly excluded.

```{r coef-compare}
p <- 100

# JFM: variables of interest (true signal or selected)
show_jfm <- sort(which(dat_jfm$alpha_true != 0 | cv_jfm$alpha != 0 |
                       dat_jfm$beta_true  != 0 | cv_jfm$beta  != 0))

coef_df <- data.frame(
  variable   = paste0("x", show_jfm),
  true_alpha = round(dat_jfm$alpha_true[show_jfm], 3),
  est_alpha  = round(cv_jfm$alpha[show_jfm],       3),
  true_beta  = round(dat_jfm$beta_true[show_jfm],  3),
  est_beta   = round(cv_jfm$beta[show_jfm],        3)
)
colnames(coef_df) <- c("variable", "alpha_true", "alpha_est", "beta_true", "beta_est")
print(coef_df, row.names = FALSE)

cat(sprintf(
  "\nJFM  alpha: TP=%d FP=%d FN=%d  |  beta: TP=%d FP=%d FN=%d\n",
  sum(cv_jfm$alpha != 0 & dat_jfm$alpha_true != 0),
  sum(cv_jfm$alpha != 0 & dat_jfm$alpha_true == 0),
  sum(cv_jfm$alpha == 0 & dat_jfm$alpha_true != 0),
  sum(cv_jfm$beta  != 0 & dat_jfm$beta_true  != 0),
  sum(cv_jfm$beta  != 0 & dat_jfm$beta_true  == 0),
  sum(cv_jfm$beta  == 0 & dat_jfm$beta_true  != 0)))
```

```{r coef-compare-jscm}
show_jscm <- sort(which(dat_jscm$alpha_true != 0 | cv_jscm$alpha != 0 |
                        dat_jscm$beta_true  != 0 | cv_jscm$beta  != 0))

coef_jscm <- data.frame(
  variable   = paste0("x", show_jscm),
  true_alpha = round(dat_jscm$alpha_true[show_jscm], 3),
  est_alpha  = round(cv_jscm$alpha[show_jscm],        3),
  true_beta  = round(dat_jscm$beta_true[show_jscm],  3),
  est_beta   = round(cv_jscm$beta[show_jscm],         3)
)
colnames(coef_jscm) <- c("variable", "alpha_true", "alpha_est", "beta_true", "beta_est")
print(coef_jscm, row.names = FALSE)

cat(sprintf(
  "\nJSCM alpha: TP=%d FP=%d FN=%d  |  beta: TP=%d FP=%d FN=%d\n",
  sum(cv_jscm$alpha != 0 & dat_jscm$alpha_true != 0),
  sum(cv_jscm$alpha != 0 & dat_jscm$alpha_true == 0),
  sum(cv_jscm$alpha == 0 & dat_jscm$alpha_true != 0),
  sum(cv_jscm$beta  != 0 & dat_jscm$beta_true  != 0),
  sum(cv_jscm$beta  != 0 & dat_jscm$beta_true  == 0),
  sum(cv_jscm$beta  == 0 & dat_jscm$beta_true  != 0)))
```

### 12.2 Time-varying AUC

We use the `timeROC` package (Blanche et al., 2013) to compute
cause-specific time-varying AUC in the competing-risk framework.
Each subject contributes at most a first-readmission event (cause 1)
and a death event (cause 2).  Each sub-model is assessed with its own
linear predictor: $\hat\alpha^\top z_i$ for readmission,
$\hat\beta^\top z_i$ for death.

> **Note**: AUC is evaluated on the training data for illustration.
> In practice use held-out or cross-validated predictions.

```{r auc-prep}
# Construct competing-risk dataset:
# Keep first readmission (event==1 & t.start==0) + death/censor (event==0).
# Status: 1 = first readmission, 2 = death, 0 = censored.
.cr_data <- function(Data) {
  d3 <- Data[Data$event == 0 | (Data$event == 1 & Data$t.start == 0), ]
  d3 <- d3[order(d3$id, d3$t.start, d3$t.stop), ]
  status <- ifelse(d3$event == 1 & d3$status == 0, 1L,
             ifelse(d3$event == 0 & d3$status == 0, 0L, 2L))
  list(data = d3, status = status)
}

cr_jfm  <- .cr_data(Data_jfm)
cr_jscm <- .cr_data(Data_jscm)

# Baseline covariates (one row per subject)
Z_jfm  <- as.matrix(Data_jfm[!duplicated(Data_jfm$id),   paste0("x", 1:p)])
Z_jscm <- as.matrix(Data_jscm[!duplicated(Data_jscm$id), paste0("x", 1:p)])

# Markers expanded to row level: alpha^T z for readmission, beta^T z for death
M_re_jfm  <- drop(Z_jfm  %*% cv_jfm$alpha)[cr_jfm$data$id]
M_de_jfm  <- drop(Z_jfm  %*% cv_jfm$beta)[cr_jfm$data$id]
M_re_jscm <- drop(Z_jscm %*% cv_jscm$alpha)[cr_jscm$data$id]
M_de_jscm <- drop(Z_jscm %*% cv_jscm$beta)[cr_jscm$data$id]
```

```{r auc, cache = TRUE}
if (!requireNamespace("timeROC", quietly = TRUE))
  install.packages("timeROC")
library(survival)
library(timeROC)

# Evaluation grid: 20 points spanning the 10th-85th percentile of event times
.tgrid <- function(t_vec, status, n = 20) {
  t_ev <- t_vec[status > 0]
  seq(quantile(t_ev, 0.10), quantile(t_ev, 0.85), length.out = n)
}

t_jfm  <- .tgrid(cr_jfm$data$t.stop,  cr_jfm$status)
t_jscm <- .tgrid(cr_jscm$data$t.stop, cr_jscm$status)

# Readmission AUC: alpha^T z marker, cause = 1
roc_re_jfm <- timeROC(T = cr_jfm$data$t.stop, delta = cr_jfm$status,
                       marker = M_re_jfm, cause = 1, weighting = "marginal",
                       times = t_jfm, ROC = FALSE, iid = FALSE)
roc_re_jscm <- timeROC(T = cr_jscm$data$t.stop, delta = cr_jscm$status,
                        marker = M_re_jscm, cause = 1, weighting = "marginal",
                        times = t_jscm, ROC = FALSE, iid = FALSE)

# Death AUC: beta^T z marker, cause = 2
roc_de_jfm <- timeROC(T = cr_jfm$data$t.stop, delta = cr_jfm$status,
                       marker = M_de_jfm, cause = 2, weighting = "marginal",
                       times = t_jfm, ROC = FALSE, iid = FALSE)
roc_de_jscm <- timeROC(T = cr_jscm$data$t.stop, delta = cr_jscm$status,
                        marker = M_de_jscm, cause = 2, weighting = "marginal",
                        times = t_jscm, ROC = FALSE, iid = FALSE)
```

```{r auc-plot, fig.height = 5}
.get_auc <- function(roc, cause) {
  auc <- roc[[paste0("AUC_", cause)]]
  if (is.null(auc)) auc <- roc$AUC
  if (is.null(auc) || !is.numeric(auc)) return(rep(NA_real_, length(roc$times)))
  if (length(auc) == length(roc$times) + 1) auc <- auc[-1]
  as.numeric(auc)
}

old_par <- par(mfrow = c(1, 2), mar = c(4.5, 4, 3, 1))

plot(t_jfm, .get_auc(roc_re_jfm, 1), type = "l", lwd = 2, col = "steelblue",
     xlab = "Time", ylab = "AUC(t)", main = "JFM", ylim = c(0.4, 1))
lines(t_jfm, .get_auc(roc_de_jfm, 2), lwd = 2, col = "tomato", lty = 2)
abline(h = 0.5, lty = 3, col = "grey60")
legend("bottomleft", c("Readmission", "Death"),
       col = c("steelblue", "tomato"), lwd = 2, lty = c(1, 2),
       bty = "n", cex = 0.85)

plot(t_jscm, .get_auc(roc_re_jscm, 1), type = "l", lwd = 2, col = "steelblue",
     xlab = "Time", ylab = "AUC(t)", main = "JSCM", ylim = c(0.4, 1))
lines(t_jscm, .get_auc(roc_de_jscm, 2), lwd = 2, col = "tomato", lty = 2)
abline(h = 0.5, lty = 3, col = "grey60")
legend("bottomleft", c("Readmission", "Death"),
       col = c("steelblue", "tomato"), lwd = 2, lty = c(1, 2),
       bty = "n", cex = 0.85)

par(old_par)
```

---

## Package conventions

- **`theta`** is always stored as `c(alpha, beta)`:
  `theta[1:p]` = `alpha` (readmission), `theta[(p+1):(2p)]` = `beta` (death).
  This holds for both the JFM and JSCM.
- `stagewise_fit()` also returns `alpha` and `beta` as separate *p* × (*k*+1)
  matrices for convenient access to the full regularization path.
- `cv_stagewise()` stores `alpha` and `beta` at the CV-optimal lambda, plus
  variable-count vectors `n_active_alpha`, `n_active_beta`, `n_active`.
- **Baseline hazards**: stored in `cv$baseline` and accessible via
  `baseline_hazard(cv, times)`.  Works for both models: Breslow estimator
  for JFM; Nelson-Aalen on the accelerated time scale for JSCM.
  Used internally by `predict()`.
- **Survival prediction**: `predict(cv, newdata)` works for both models and
  returns a `swjm_pred` object with `S_re`, `S_de` (survival matrices),
  linear predictors `lp_re`, `lp_de`, and predictor-contribution matrices
  `contrib_re`, `contrib_de` ($\hat\alpha_j z_{ij}$ and $\hat\beta_j z_{ij}$).
  - **JFM**: contributions are log-hazard-ratio contributions (positive =
    higher risk).
  - **JSCM**: also returns `time_accel_re` and `time_accel_de`
    ($e^{\hat\alpha^\top z_i}$, $e^{\hat\beta^\top z_i}$) — the multiplicative
    factors by which each subject's event times are scaled relative to baseline.
    Contributions $\hat\alpha_j z_{ij}$ are log time-acceleration contributions:
    $e^{\hat\alpha_j z_{ij}} > 1$ shortens event times; $< 1$ lengthens them.
- The **cooperative lasso** norm penalizes each variable pair
  `(alpha_j, beta_j)` with the L2 norm when signs agree, and L1 when they
  disagree — encouraging variables that affect both processes to enter
  together with the same sign.
- **Gradient scaling**: at each step the death gradient is scaled up so that
  both sub-models contribute comparably to the penalty norm.
- **Adaptive step size**: the step size at each iteration is set to one order
  of magnitude smaller than the current dual norm, which avoids overly large
  updates.
